\textls[-0]{\thetitle}

### 2023-03-13 metatheory2.tex
Fix issue on variables.
Save. Adding universe.

In this section, we lay out the complete version of the syntax of \TT but in the style of \citeauthor{debruijn1964} indices and \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., in the form of \sub{T}{\gm}) are part
of the syntax rather than meta-operations.

@@

\subsection{Review MLTT in Debruijn Indices and Explicit Subst}

@@

The change of the syntax formulation is important to the simplicity of our proof, as we want to handle the proof about substitutions and terms uniformly---thus making substitution as part of the syntax (explicit substitution) will be helpful.

Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

>>>>

In this section, we lay out the complete version of the syntax of \TT but in the style of \citeauthor{debruijn1964} indices and \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., in the form of \sub{T}{\gm}) are part
of the syntax rather than meta-operations. What's more, we also make universe level explicit for completeness.

@@

\subsection{Review MLTT in Debruijn Indices, Explicit Subst, and Explicit Universe Level}

@@

\noindentparagraph{Debruijn Indices and Explicit Substitution.}  The change of the syntax formulation is important to the simplicity of our proof, as we want to handle the proof about substitutions and terms uniformly---thus making substitution as part of the syntax (explicit substitution) will be helpful.

Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

Finally, $\SubstWeak{n}$ is a short hand for $\pi_1^n~"id"$ and $\var{n}$ is a short hand for $\pi_2~\pi_1^{n}$. Thus during meta-theoretic reasoning, we will only deal with $\pi_1$ and $\pi_2$.

\noindentparagraph{Explicit Universe Level.} Our treatment of universe level mainly follows that of \citet{altkap2016, kaposi2019gluing}. Universe level is mainly to get around the size issue---we cannot refer to(and thus quantifiy over) a set of all set (or a universe of all type). So everytime we want to quantify over a universe of types, we have to specify ``how large'' that universe is, that is basically what universe level is measuring.

% We first need the mathmatical setup to talk about universe level, following \citet{sterling2019algebraic}:

% \newcommand{\Set}[1]{\ensuremath{\texttt{Set}_{#1}}}

% {\textbf{Set-theoretic Universe Assumption.}} As a simplifying move, we assume an infinite hierarchy of Grothendieck universes $\Set{i}$ for $i \in \mathbb{N}$ in our meta-logic.

% For people familiar with Agda, we can consider $\Set{i}$ as the one in Agda. In other words, each $\Set{i}$ is (1). closed under dependent function type and dependent tuple type, (2). cumulative as $\Set{i} \in \Set{i+1}$ and $\Set{i} \subseteq \Set{i+1}$

Now we will need to attach syntax pieces with natural number indicating universe level. The explicit universe level $i$ now will attach to each context and type judgement $\goodCtx{\Gm}{i}$ and $\goodType{\Gm}{T}{i}$ basically indicating the context $\Gm$ and $T$ inhabits at what universe. Apparently, universe level will also attach to $\cU$. For two universe level $i,j$ (as natural number), we use $i \lcup j$ to denotes $\max(i,j)$.
